Breaking Down the NFL: How Offensive Yards Shape Game Outcomes

### 2024-04-05 paper/paper.qmd
update paper

---
title: "Field Dynamics" 
subtitle: "Why the Rush Game is still King in the Modern Day NFL"
author: "Alexander Guarasci"
thanks: "Code and data underpinning this paper are available at: https://github.com/AlexanderG123/nfl_analysis"
date: "`r format(Sys.Date(), '%B %d, %Y')`" # This will add the current date
date-format: long
bibliography: references.bib
execute:
  echo: false
format: pdf
fig-pos: "H"
abstract: |
 This study examines the strategy behind National Football League (NFL) games, focusing on how teams' offensive yardage—gained through both rushing and passing—affects their chances of winning. Analyzing play-by-play data from the 2023 NFL season, we discovered that while both rushing and passing yards contribute to game victories, a stronger rushing attack significantly increases a team's likelihood of winning. These findings highlight the critical role of offensive strategy in football, providing valuable insights for teams looking to optimize their game plans for better outcomes.

@@

# Introduction {#sec-intro}

@@

This comprehensive dataset encapsulates a wide array of variables, central to which are “RushingYards”, “PassingYards”, “TotalYards”, “home_score”, “away_score”, “Winner”, “Win”, “Q4_RushingYards”, “Q4_PassingYards” and “Q4_TotalYards”. RushingYards and  PassingYards represent the yards gained by a team through rushing and passing plays, respectively, within a single game, while Q4_RushingYards and Q4_PassingYards represent the rushing and passing yards in the fourth-quarter. “RushingYards” is taken as the sum of the variable “rushing_yards” from the original dataset with the same “game_id” (so that we get the rushing yards from each game) and “posteam” (so each team that plays gets has their rushing yards tracked).  PassingYards, Q4_RushingYards and Q4_PassingYards are similarly constructed. TotalYards is a constructed variable, representing the sum of RushingYards and PassingYards, devised to encapsulate a team's overall offensive performance. Q4_TotalYards is constructed as the sum of Q4_RushingYards and Q4_PassingYards. The scores of the home and away teams are captured by home_score and away_score, respectively, which is data that is provided and that does not need to be constructed. The Winner is determined based on the comparison of these scores, classified as the team that won (for example, “Bal” if the Baltimore Ravens won). Finally, the binary Win variable signifies the outcome from the perspective of the team being analyzed (1 for a win, 0 for a loss, and “TIE” for a tie, but because there were no ties last season there are no instances of “TIE”).

@@

The first logistic regression model scrutinizes the entire game, using rushing and passing yards as predictor variables to estimate the odds of winning. This model, expressed as $$
\log\left(\frac{p_1}{1 - p_1}\right) = \beta_0 + \beta_1 \cdot \text{RushingYards} + \beta_2 \cdot \text{PassingYards}
$$
, where p1 denotes the probability of winning, facilitates a broad analysis of strategic effectiveness. The coefficients, representing rushing and passing yards respectively, illuminate the value of each yard gained in the context of winning. The positive coefficient for rushing yards implies a direct correlation between a stronger rushing offence and the likelihood of victory, underscoring the strategic advantage of a balanced offensive game. This model is apt for providing a holistic view of game strategy, allowing analysts and teams to gauge the overall impact of their offensive efforts throughout the match.

@@

The second logistic regression model narrows the focus to the fourth quarter, a critical period where games are often decided. By analyzing the same predictors but confined to the final quarter, this model, $$
\log\left(\frac{p_2}{1-p_2}\right) = \beta_3 + \beta_4 \cdot \text{Q4\_RushingYards} + \beta_5 \cdot \text{Q4\_PassingYards}
$$
, offers insights into the tactical shifts that dominate the endgame. Unlike the overall game model, the fourth-quarter model might reveal a nuanced strategic landscape, such as a negative coefficient for passing yards, indicating the potential pitfalls of a pass-heavy approach under pressure. This focused analysis is instrumental in understanding the dynamics of clutch gameplay, shedding light on the adjustments teams make in response to the scoreboard and the clock.

>>>>

---
title: "Breaking Down the NFL: How Offensive Yards Shape Game Outcomes"
subtitle: "Examining the Trends in Yardage of Winning NFL Teams Using Logistic Regression"

author: "Alexander Guarasci"
thanks: "Code and data underpinning this paper are available at: https://github.com/AlexanderG123/nfl_analysis"
date: "`r format(Sys.Date(), '%B %d, %Y')`" # This will add the current date
date-format: long
bibliography: references.bib
execute:
  echo: false
format: pdf
fig-pos: "H"
abstract: |
 This study examines the strategy behind National Football League (NFL) games, focusing on how teams' offensive yardage—gained through both rushing and passing—affects their chances of winning using two logistic regression models. Analyzing play-by-play data from the 2023 NFL season, we discovered that while both rushing and passing yards contribute to game victories, a stronger rushing attack significantly increases a team's likelihood of winning. These findings highlight the critical role of offensive strategy in football, providing valuable insights for teams looking to optimize their game plans for better outcomes.

@@

# Introduction {#sec-introduction}

@@

This comprehensive dataset encapsulates a wide array of variables, central to which are “**RushingYards**”, “**PassingYards**”, “**TotalYards**”, “**home_score**”, “**away_score**”, “**Winner**”, “**Win**”, “**Q4_RushingYards**”, “**Q4_PassingYards**” and “**Q4_TotalYards**”. **RushingYards** and  **PassingYards** represent the yards gained by a team through rushing and passing plays, respectively, within a single game, while **Q4_RushingYards** and **Q4_PassingYards** represent the rushing and passing yards in the fourth-quarter. **RushingYards** is taken as the sum of the variable “**rushing_yards**” from the original dataset with the same “**game_id**” (so that we get the rushing yards from each game) and “**posteam**” (so each team that plays gets has their rushing yards tracked).  **PassingYards**, **Q4_RushingYards** and **Q4_PassingYards** are similarly constructed. **TotalYards** is a constructed variable, representing the sum of **RushingYards** and **PassingYards**, devised to encapsulate a team's overall offensive performance. **Q4_TotalYards** is constructed as the sum of **Q4_RushingYards** and **Q4_PassingYards**. The scores of the home and away teams are captured by **home_score** and **away_score**, respectively, which is data that is provided and that does not need to be constructed. The **Winner** is determined based on the comparison of these scores, classified as the team that won (for example, “Bal” if the Baltimore Ravens won). Finally, the binary **Win** variable signifies the outcome from the perspective of the team being analyzed (1 for a win, 0 for a loss, and “TIE” for a tie, but because there were no ties last season there are no instances of “TIE”).

@@

The first logistic regression model scrutinizes the entire game, using rushing and passing yards as predictor variables to estimate the odds of winning. This first model, @eq-model1, expressed as $$
\log\left(\frac{p_1}{1 - p_1}\right) = \beta_0 + \beta_1 \cdot \text{RushingYards} + \beta_2 \cdot \text{PassingYards}
$$ {#eq-model1}
, where p1 denotes the probability of winning, facilitates a broad analysis of strategic effectiveness. The coefficients, representing rushing and passing yards respectively, illuminate the value of each yard gained in the context of winning. The positive coefficient for rushing yards implies a direct correlation between a stronger rushing offence and the likelihood of victory, underscoring the strategic advantage of a balanced offensive game. This model is apt for providing a holistic view of game strategy, allowing analysts and teams to gauge the overall impact of their offensive efforts throughout the match.

@@

The second logistic regression model, @eq-model2, narrows the focus to the fourth quarter, a critical period where games are often decided. By analyzing the same predictors but confined to the final quarter, this model, $$
\log\left(\frac{p_2}{1-p_2}\right) = \beta_3 + \beta_4 \cdot \text{Q4\_RushingYards} + \beta_5 \cdot \text{Q4\_PassingYards}
$$ {#eq-model2}
, offers insights into the tactical shifts that dominate the endgame. Unlike the overall game model, the fourth-quarter model might reveal a nuanced strategic landscape, such as a negative coefficient for passing yards, indicating the potential pitfalls of a pass-heavy approach under pressure. This focused analysis is instrumental in understanding the dynamics of clutch gameplay, shedding light on the adjustments teams make in response to the scoreboard and the clock.
